---
title: "scRNAseq QC and Processing"
author: "Me!"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
## If you don't have these installed, running the helper script 
## "install_packages-02.R" can quickly install what you don't have
library(Seurat)      
library(tidyverse)   ## popular R language extension that includes piping and ggplot
library(here)        ## allows you to refer to relative paths based on projects
library(clustree)    ## visualize clustering resolutions
library(scDblFinder) ## find doublets

knitr::opts_chunk$set(echo = FALSE,
                      warning = TRUE,
                      message = TRUE,
                      include = TRUE,
                      fig.width = 8, fig.height = 8,
                      fig.align = 'center',
                      cache = FALSE, cache.lazy = FALSE)

## Explicitly setting the randomizer seed for reproducibility 
set.seed(43)
## Sourcing a small set of custom functions as a separate script to keep this
## script shorter and more readable
source(here('Session-02/functions_session-02.R'))
```

# Load data

Reading in CellRanger outputs

```{r}
obj <- Read10X_h5(here('Data/kang/s1244_stim.h5'))
obj <- CreateSeuratObject(obj)
```

Initializing a list to record changes to our dataset through processing steps

```{r}
qc_receipts <- list()
qc_receipts$raw_data <- dim(obj)
```

# Add QC metrics

We can write new metadata to our Seurat object

```{r}
# Add number of genes per UMI for each cell to metadata
obj$log10GenesPerUMI <- log10(obj$nFeature_RNA)/log10(obj$nCount_RNA)

## Detecting features based on pattern recognition through regular expressions
## If you're not familiar with regex, you can read more here
## https://en.wikipedia.org/wiki/Regular_expression

# Mitochondrial ratio
obj$percent.mt <- PercentageFeatureSet(object = obj, pattern = "^MT")/100

# Ribosomal ratio
obj$percent.rb <- PercentageFeatureSet(object = obj, pattern = "^RP[SL]")/100
```

Seurat has some built in visualization functions

```{r}
VlnPlot(obj, features = c('nFeature_RNA', 'nCount_RNA', 'percent.mt'), ncol = 3)
```

```{r}
FeatureScatter(obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + 
  aes(alpha = 0.5) + 
  NoLegend()
```

We can make custom plots of the metadata 

```{r}
ggplot(obj@meta.data, aes(x=log10GenesPerUMI, y=nCount_RNA, color=percent.mt)) +
  geom_point(alpha=0.5) +
  scale_color_gradient(low = 'goldenrod1', high = 'red', limits = c(0,1)) +
  scale_x_log10() +
  scale_y_log10() +
  theme_classic()
```

# Identify doublets

This method uses [scDblFinder](https://plger.github.io/scDblFinder/index.html).
Other packages such as [DoubletFinder](https://github.com/chris-mcginnis-ucsf/DoubletFinder)
perform similar processing.

```{r}
## Custom function, see 'R/functions_session-02.R' for more details
obj <- run_scDblFinder(obj)
```

You can subset the seurat object using metadata columns and values

```{r}
## keeping only the singlets
obj <- subset(obj, subset = scDblFinder.class == 'singlet')
## observe how many cells were removed
qc_receipts$post_doublet_removal <- dim(obj)
qc_receipts
```

# Filtering low quality cells

## Set filtering thresholds

Using an outlier detection approach to initialize QC cutoff thresholds. 
You could also manually specify your filtering thresholds. Reasonable
cutoffs depend on the source material and study goals, so tailor
these as necessary. 

```{r}
## Custom function, see 'R/functions_session-02.R' for more details
## nmads = number of median absolute devations used for outlier threshold
cutoffs <- generate_capture_QC_cutoffs(obj, nmads = 5)
cutoffs
```

Manually adjust thresholds here if necessary

```{r}
cutoffs$nCount_RNA.max <- 10000
cutoffs$percent.mt.max <- 5
```

Now adding metadata to identify which cells fall within thresholds

```{r}
obj$filtered <- 
  !(obj$nCount_RNA >= cutoffs$nCount_RNA.min &
    obj$nCount_RNA <= cutoffs$nCount_RNA.max &
    obj$nFeature_RNA >= cutoffs$nFeature_RNA.min &
    obj$nFeature_RNA <= cutoffs$nFeature_RNA.max &
    obj$log10GenesPerUMI >= cutoffs$log10GenesPerUMI.min &
    obj$log10GenesPerUMI <= cutoffs$log10GenesPerUMI.max &
    obj$percent.mt >= cutoffs$percent.mt.min &
    obj$percent.mt <= cutoffs$percent.mt.max &
    obj$percent.rb >= cutoffs$percent.rb.min &
    obj$percent.rb <= cutoffs$percent.rb.max)
```

## Visualize outliers

```{r}
ggplot(obj@meta.data, aes(x=nCount_RNA, y=nFeature_RNA, color=filtered)) +
  geom_point(alpha=0.5) +
  scale_color_manual(values = c('black', 'orange')) +
  scale_x_log10() +
  scale_y_log10() +
  theme_classic()
```

## Filter cells

```{r}
obj <- subset(obj, subset = filtered == FALSE)
qc_receipts$post_filtering <- dim(obj)
```

# QC filtering summary

```{r}
qc_receipts
```


# Normalize and scale

```{r}
## Seurat functions have lots of arguments with default values, 
## consider checking the help documentation to learn more about these functions
obj <- NormalizeData(obj, normalization.method = 'LogNormalize')
obj <- ScaleData(obj)
```

Adjusted data are stored in new slots in the Seurat object

```{r}
## Normalized data
obj@assays$RNA@data[c('CD3D'),1:6]
## Scaled data
obj@assays$RNA@scale.data[c('CD3D'),1:6]
```

# Dimensional reduction

## Find variable features

Focus on the most variable genes in the dataset. These will be the most informative
for separating cell populations.

```{r}
obj <- FindVariableFeatures(obj)
## Variable features are stored within the assay
head(obj@assays$RNA@var.features)
```

## Calculate PCs

Seurat's PCA function uses the variable features defined earlier. 

```{r}
obj <- RunPCA(obj)
```

You can see how much a given gene is influencing a principal component

```{r}
obj@reductions$pca@feature.loadings[c('CCR7', 'CCL8', 'ISG15'), 1:5]
```

# Visualization

Visualization with PCA doesn't show distinct cell populations. We'll use UMAP
to visualize our data

```{r}
## This is supposed to look bad
DimPlot(obj, group.by = 'orig.ident', reduction = 'pca') + NoLegend()
```

## Identify relevant PCs

Don't get too hung up on this: most state-of-the-field methods
are pretty robust to PC selection. The most important point is to 
include all of the most high-variance PCs.

```{r}
ElbowPlot(obj, ndims = 30)
```

## Construct neighbor graph

Identify neighbors in a lower-dimension space. Neighbors will be used to inform clusters later, but we do it here since it depends on significant PCs.

```{r}
obj <- FindNeighbors(obj, dims = 1:10)
obj@graphs$RNA_snn[1:6, 1:6]
```

## Calculating UMAP

Recall that UMAP is a visualization technique to reduce high-dimensional data into lower space for plotting. We perform UMAP on the significant principle components.

```{r}
obj <- RunUMAP(obj, dims = 1:10)
```

The result is a visual separation of cells based on highly variable features
captured in the PCs.

```{r}
DimPlot(obj, group.by = 'orig.ident', reduction = 'umap') + NoLegend()
```

We often use the UMAP for visualizing features and metadata on cell populations 
in a capture. But ultimately the UMAP output is just coordinate values for plotting. 

```{r}
head(obj@reductions$umap@cell.embeddings)
```

# Capture level metadata clustering

Check if any relevant metadata is clustering in interesting ways. For example, a 
cluster of high mitochndrial expression may indicate dead or dying cells that you 
may want to exclude from analysis.

```{r, fig.width=12, fig.height=12}
## Running `RunUMAP` usuaslly sets your default reduction to umap
FeaturePlot(obj,
            features = c('nFeature_RNA',
                         'nCount_RNA',
                         'percent.mt',
                         'percent.rb', 
                         'log10GenesPerUMI',
                         'CD3G'))
```

# Check cell cycle

Seurat provides some pre-defined cell-cycle genes. 
See if any CC genes are in your variable features. 
If not, you probably don't need to do this.

```{r}
c(cc.genes$s.genes, cc.genes$g2m.genes)[c(cc.genes$s.genes, cc.genes$g2m.genes) %in% VariableFeatures(obj)]
```

It's also useful to look at your final clustering to see if it's actually
being driven by cell cycle. Look for separation of phases. 

```{r}
## Seurat has functions to calculate a module score, weighting
## the expression of many genes into a single value. 
## Seurat provides a list of cell-cycle associated genes that we pass
## to their function. The output includes cell cycle scores 
## and a single phase classification.
obj <- CellCycleScoring(obj,
                        s.features = cc.genes$s.genes,
                        g2m.features = cc.genes$g2m.genes)
DimPlot(obj, group.by = 'Phase')
```

If you see something that looks like it's being driven by cell cycle,
you can attempt to regress the expression signature associated with cell cycle
out from your data.

This can take a while. To quickly inspect if cell cycle regression will be fruitful, 
we will focus on the subset of genes present in more than N cells. 

```{r}
## Custom function, see 'R/functions_session-02.R' for more details
obj.cc_check <- filterLowGenes(obj, min.cells = 100, assay = 'RNA', calculate_only = FALSE)
```

```{r}
obj.cc_check <- ScaleData(obj.cc_check,
                          vars.to.regress = c('S.Score', 'G2M.Score'),
                          features = rownames(obj.cc_check)) %>%
  FindVariableFeatures() %>%
  RunPCA() %>%
  FindNeighbors(dims = 1:20) %>%
  RunUMAP(dims = 1:20)
  
DimPlot(obj.cc_check, reduction = 'umap', group.by = 'Phase')
```

If this doesn't look like an improvement, but you still want to regress cell cycle
from the data, you can try an alternative workflow. See Seurat's vignette for more details.
https://satijalab.org/seurat/articles/cell_cycle_vignette

```{r}
obj.cc_check$CC.difference <- obj.cc_check$S.Score - obj.cc_check$G2M.Score

obj.cc_diff_check <- ScaleData(obj.cc_check,
                          vars.to.regress = c('CC.difference'),
                          features = rownames(obj.cc_check)) %>%
  FindVariableFeatures() %>%
  RunPCA() %>%
  FindNeighbors(dims = 1:20) %>%
  RunUMAP(dims = 1:20)
  
DimPlot(obj.cc_diff_check, reduction = 'umap', group.by = 'Phase')
```

# Clustering

The clusters are informed by the neighbors graph we constructed earlier in the 
dimensional reduction section. There are many parameters to tweak in defining 
clusters and creating neighbor graphs, but the most commonly altered one is the 
resolution. High resolution generally produces many smaller clusters, while lower resolutions generally produce few large clusters. A cluster tree can help 
visualize changes in clusters across different resolutions.

You can also pull out subsets of your data and independently cluster those, force
higher-resolution clustering of particular clusters you wish to break up, or join 
clusters with no observable distinction. Cluster defining is often an iterative 
process as you work to identify the cell populations in your data.

```{r}
## Making a temporary copy of the object so we don't have all this
## unneeded data added to our working object
clustree_data <- obj
for (i in seq(0.2,1.4,0.2)){
  clustree_data <- FindClusters(clustree_data, resolution = i, verbose = FALSE)
}
```

Look for a resolution that is stable: no strong changes in cluster membership
at either higher or lower resolutions. 

```{r, fig.height=12, fig.width=12}
clustree(clustree_data)
```

```{r}
obj <- FindClusters(obj, resolution = 0.8)
```

We can now project our clusters onto the UMAP. Running 'FindClusters' automatically
sets the identity of cells in our object as the cluster ID, so no additional arguments
are needed.

```{r}
DimPlot(obj, label = TRUE)
```

# Save processed object

```{r}
saveRDS(obj, here('saved_rds/processed_obj.Rds'))
```

