---
title: "scRNAseq Markers and Cell annotation"
author: "Me!"
date: "`r Sys.Date()`"
output: html_document
---

# 0 - Project Setup

## Load packages

```{r}
library(Seurat)
library(UCell)
library(SingleR)
library(celldex)
```

## Downloading the necessary files

Make sure that you've downloaded a zip of the github repo [INSERT URL HERE] and unzip it somewhere on your computer (e.g. in your `Desktop` directory).

If you want to keep working in the directory you created last class, simple copy and paste the `04_markers+annotation` folder into the base level of your existing directory.

If you'd like to start fresh in a clean working directory, you can simply work from the newly unzipped folder.

*Note: If you are comfortable using git (beyond the scope of this workshop) and are already working within a local clone of the repo, you can skip the above and simply `git pull` the repo.*

## Setting up an RStudio Project and using here()

If you aren't already working within an RStudio Project with the `scRNAseqWorkshop2023` folder as the base level, create a new Project:
  `File` (in the RStudio toolbar at the top of your screen) > 
    `New Project...` (in the dropdown menu) > 
      `Existing Directory` (in the prompt that appears) > 
        Navigate to the base level of the directory that you just unzipped (likely named `scRNAseqWorkshop2023`) >
          OK

Now, in the R Console (or in this chunk) run `getwd()`. It should be the full path to the `scRNAseqWorkshop2023` directory.
```{r}
getwd()
```

The `here()` function from the `here` package (`install.packages("here")` if you didn't last week) automatically knows the base level of your RStudio project so that you can refer to files using their relative paths. This makes it easier to share a project with others (or yourself on another machine) without needing to hard-code absolute paths. Check that it is working by running the following:
```{r}
# should output the same path as getwd()
library(here)

# if that didn't print a message run this
here()

# this function call automatically tests whether the correct directory is assigned to here(), and attempts to fix it if not
i_am("04_markers+annotation/04-markers+annotation_script.Rmd")

# try this again just in case
here()
```

Now that you've set up `here()` correctly, don't try to change directories using `setwd()`. Just refer to files based on their relative paths starting at the project's base directory within the `here()` function.

***

**Add section reviewing the UMAP and unsupervised clusters that we start with.**

# 1 - Using *a priori* Markers to identify cell types

## 1.1 - Visualize individual markers

### FeaturePlot

```{r}

```

### VlnPlot

```{r}

```

### DotPlot

```{r}

```


## 1.2 - Visualize group signatures

### Calculate UCell module score for each cell type in the marker list

```{r}

```

### Plot FeaturePlots split across each label

```{r}

```

### Plot a DotPlot for each module score

```{r}

```


***

# 2 - Finding *de novo* Markers

## 2.1 - Find one-vs-all markers for each unsupervised cluster

```{r}

```

## 2.2 - Plot heatmap of all markers

```{r}

```

## 2.3 - Custom GO analysis dotplot

```{r}
# refer to helper functions script
```

***

# 3 - Automated cell type annotation methods


